RVD2: An ultra-sensitive variant detection model for heterogeneous population sequencing

### 2013-09-05 doc/bioinformatics2013/RVD2.tex
fixed appendix title
Refined the Initialization section and its derivation.

\begin{figure}[h]
\begin{center}
\includegraphics[width=40mm]{pdf_figs/RVD2_model.pdf}
\caption{RVD2 Graphical Model}
\label{fig:graphical_model}
\end{center}
\end{figure}

@@

The estimated error probability is calculated from the observed sample
\begin{equation}
 \tilde{\theta}_{ij}=\frac{r_{ij}}{n_{ij}}
\end{equation}
The estimate of parameter $\mu_j$ is obtained using the expected value of $\theta_{ij}$, while $M_j$ is estimated using the moment estimates for the variance of $\theta_{ij}$.
\begin{equation}
 \mu_j=E(\theta_{ij}|\mu_j,M_j)
\end{equation}
\begin{equation}
 \tilde{\mu}_j =\frac{1}{N-1}\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}
\end{equation}

\begin{equation}
 M_j=\frac{\mu_j(1-\mu_j)}{Var(\theta_{ij}|\mu_j,M_j)}-1
\end{equation}
\begin{equation}
 \tilde{M}_j = \frac{\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\cdot\left(1-\frac{1}{N-1}\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\right)}{\sum_{i=1}^N\left(\frac{r_{ji}}{n_{ji}}-\frac{1}{N-1}\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\right)^2}
\end{equation}
Likewise,the estimate of hyperparameters $\mu_o$ and $M_0$ are
\begin{equation}
 \mu_0=E(\mu_j|\mu_0,M_0)
\end{equation}
\begin{equation}
 \tilde{\mu}_0 =\frac{1}{N-1}\sum_{i=1}^N\tilde{\mu}_j
\end{equation}

\begin{equation}
 M_0=\frac{\mu_0(1-\mu_0)}{Var(\mu_j|\mu_0,M_0)}-1
\end{equation}
\begin{equation}
 \tilde{M}_0 = \frac{\sum_{i=1}^J\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\cdot\left(1-\frac{1}{(J-1)(N-1)}\sum_{i=1}^J\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\right)}{\sum_{i=1}^J\left(\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}-\frac{1}{J-1}\sum_{i=1}^J\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\right)^2}-1
\end{equation}

@@

The null hypothesis is tested using power-divergence family of statistics proposed by Cressie and Read (1984). Considering observed frequencies $\{r_i; i=1,...,k\}$ and expected frequencies $\{E_i; i=1,...,k\}$, by definition the power-divergence family of statistcs indexed by real parameter $\lambda$ is 
\begin{equation}
 2nI^\lambda = \frac{2}{\lambda(\lambda+1)}\sum_{i=1}^k r_i \left[\left(\frac{r_i}{E_i}\right)^\lambda-1\right];\lambda \in R
\end{equation}

@@

\section{Appendix}
\subsection{Parameter initialization derivation}
The model parameters and latent variables are initialized using method of moments by equating population moment to sample moment.

The estimated error probability is calculated from the observed sample
\subsubsection{$\theta_{ji}$}\
\begin{equation}
 \tilde{\theta}_{ji}=\frac{r_{ji}}{n_{ji}}
\end{equation}

\subsubsection{$\mu_j$ and $M_j$}\
\begin{align}
 \theta_{ji}|\mu_j,M_j\thicksim \text{Beta}(\mu_j,M_j)
\end{align}

@@

The second population moment centered around mean (variance) is
\begin{equation}
 Var(\theta_{ji}) = \frac{\mu_j(1-\mu_j)}{M_j+1}
\end{equation}
while the sample variance is
\begin{equation}
 \begin{split}
  Var(\tilde{\theta}_{ji}) &= \frac{1}{N-1}\sum_{i=1}^N(\tilde{\theta}_{ji}-\text{E}[\tilde{\theta}_{ji}])^2\\ &= \frac{1}{N-1}\sum_{i=1}^N(\tilde{\theta}_{ji}-\tilde{\mu}_j)^2
 \end{split}
\end{equation}
By definition of Method of moments,
\begin{equation}
  \frac{\tilde{\mu}_j(1-\tilde{\mu}_j)}{\tilde{M}_j+1} = \frac{1}{N-1}\sum_{i=1}^N(\tilde{\theta}_{ji}-\tilde{\mu}_j)^2,
\end{equation}
so,
\begin{equation}
\begin{split}
 \tilde{M}_j &= (N-1)\cdot\frac{\tilde{\mu}_j(1-\tilde{\mu}_j)}{\sum_{i=1}^N(\tilde{\theta}_{ji}-\tilde{\mu}_j)^2}-1\\ &= (N-1)\cdot\frac{\frac{1}{N-1}\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\cdot\left(1-\frac{1}{N-1}\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\right)}{\sum_{i=1}^N\left(\frac{r_{ji}}{n_{ji}}-\frac{1}{N-1}\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\right)^2}\\ &= \frac{\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\cdot\left(1-\frac{1}{N-1}\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\right)}{\sum_{i=1}^N\left(\frac{r_{ji}}{n_{ji}}-\frac{1}{N-1}\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\right)^2}
\end{split}
\end{equation}

\subsubsection{$\mu_0$ and $M_0$}
\begin{align}
 \mu_j|\mu_0,M_0\thicksim \text{Beta}(\mu_0,M_0)
\end{align}
Likewise,
\begin{equation}
 \tilde{\mu}_0 =\frac{1}{N-1}\sum_{i=1}^N\tilde{\mu}_j
\end{equation}

@@

Using Methond of Moment
\begin{equation}
  \tilde{\mu}_0 = \frac{1}{(J-1)(N-1)}\sum_{i=1}^J\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}
\end{equation}

The second population moment centered around mean (variance) is
\begin{equation}
 Var(\mu_j) = \frac{\mu_0(1-\mu_0)}{M_0+1}
\end{equation}
while the sample variance is
\begin{equation}
 \begin{split}
  Var(\tilde{\mu}_j) &= \frac{1}{J-1}\sum_{i=1}^J(\tilde{\mu}_j-\text{E}[\tilde{\mu}_j])^2\\ &= \frac{1}{J-1}\sum_{i=1}^J(\tilde{\mu}_j-\tilde{\mu}_0)^2
 \end{split}
\end{equation}
By definition of Method of moments,
\begin{equation}
  \frac{\tilde{\mu}_0(1-\tilde{\mu}_0)}{\tilde{M}_0+1} = \frac{1}{J-1}\sum_{i=1}^J(\tilde{\mu}_j-\tilde{\mu}_0)^2,
\end{equation}
so,
\begin{equation}
\begin{split}
 \tilde{M}_0 &= (J-1)\cdot\frac{\tilde{\mu}_0(1-\tilde{\mu}_0)}{\sum_{i=1}^J(\tilde{\mu}_j-\tilde{\mu}_0)^2}-1\\ &= (J-1)\cdot\frac{\frac{1}{(J-1)(N-1)}\sum_{i=1}^J\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\cdot\left(1-\frac{1}{(J-1)(N-1)}\sum_{i=1}^J\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\right)}{\sum_{i=1}^J\left(\frac{1}{N-1}\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}-\frac{1}{(J-1)(N-1)}\sum_{i=1}^J\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\right)^2}-1\\ &= \frac{\sum_{i=1}^J\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\cdot\left(1-\frac{1}{(J-1)(N-1)}\sum_{i=1}^J\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\right)}{\sum_{i=1}^J\left(\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}-\frac{1}{J-1}\sum_{i=1}^J\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\right)^2}-1
\end{split}
\end{equation}
\bibliographystyle{apalike}
\bibliography{bioinfo}
\end{document}

>>>>

\begin{figure}[h]
\begin{center}
%\includegraphics[width=40mm]{pdf_figs/RVD2_model.pdf}
\caption{RVD2 Graphical Model}
\label{fig:graphical_model}
\end{center}
\end{figure}

@@

The estimated error probability is calculated from the observed sample
\begin{equation}
 \tilde{\theta}_{ji}=\frac{r_{ji}}{n_{ji}}
\end{equation}

In our model, $\theta_{ji}$ is follows beta distribution with parameters $\mu_j$ and $M_j$
\begin{align}
 \theta_{ji}|\mu_j,M_j\thicksim \text{Beta}(\mu_j,M_j)
\end{align}

The first polulation moment (mean) is 
\begin{equation}
 \mu_j=E(\theta_{ij})
\end{equation}
By definition of Method of Moments, 
\begin{equation}
 \tilde{\mu}_j = E(\tilde{\theta}_{ij}) = \frac{1}{N-1}\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}
\end{equation}

The second population moment centered around mean (variance) is
\begin{equation}
 Var(\theta_{ji}) = \frac{\mu_j(1-\mu_j)}{M_j+1}
\end{equation}
while the sample variance is
\begin{equation}
 \begin{split}
  Var(\tilde{\theta}_{ji}) &= \frac{1}{N-1}\sum_{i=1}^N(\tilde{\theta}_{ji}-\text{E}[\tilde{\theta}_{ji}])^2\\ &= \frac{1}{N-1}\sum_{i=1}^N(\tilde{\theta}_{ji}-\tilde{\mu}_j)^2
 \end{split}
\end{equation}
Equating polulation variance to sample variance we have
\begin{equation}
 \begin{split}
  \tilde{M}_j &=\frac{\mu_j(1-\mu_j)}{\text{Var}(\tilde{\theta}_{ji})}-1\\ &= \frac{\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\cdot\left(1-\frac{1}{N-1}\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\right)}{\sum_{i=1}^N\left(\frac{r_{ji}}{n_{ji}}-\frac{1}{N-1}\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\right)^2}-1
 \end{split}
\end{equation}

$\mu_j$ is modeled as a Beta distribution random variable with paprameters $\mu_o$ and $M_0$. Likewise,
\begin{equation}
\begin{split}
 \tilde{\mu}_0 &= E(\mu_j)\\ &= \frac{1}{N-1}\sum_{i=1}^N\tilde{\mu}_j\\ &= \frac{1}{(J-1)(N-1)}\sum_{i=1}^J\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}
\end{split}
\end{equation}
\begin{equation}
\begin{split}
 \tilde{M}_0 &=\frac{\tilde{\mu}_0(1-\tilde{\mu}_0)}{Var(\tilde{\mu}_j)}-1\\ &= \frac{\left(\sum_{i=1}^J\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\right)\cdot\left(1-\frac{1}{(J-1)(N-1)}\sum_{i=1}^J\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\right)}{\sum_{i=1}^J\left(\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}-\frac{1}{J-1}\sum_{i=1}^J\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\right)^2}-1
\end{split}
\end{equation}

@@

The null hypothesis is tested using power-divergence family of statistics proposed by Cressie and Read (1984). By definition, the power-divergence family of statistcs parameterized with $\lambda$ is 
\begin{equation}
 2nI^\lambda = \frac{2}{\lambda(\lambda+1)}\sum_{i=1}^k r_i \left[\left(\frac{r_i}{E_i}\right)^\lambda-1\right];\lambda \in R,
\end{equation}

where $\{r_i; i=1,...,k\}$ denote observed frequencies and $\{E_i; i=1,...,k\}$ denote the expected frequencies.

@@

\section{Appendix}
\subsection{Parameter initialization derivation}\

The model parameters and latent variables are initialized using Method of Moments.
\subsubsection{$\theta_{ji}$}\

The estimated error probability is calculated from the observed sample
\begin{equation}
 \tilde{\theta}_{ji}=\frac{r_{ji}}{n_{ji}}
\end{equation}

\subsubsection{$\mu_j$ and $M_j$}\

$\theta_{ji}$ is modeled as beta distribution random variable with parameter $\mu_j$ and $M_j$
\begin{align}
 \theta_{ji}|\mu_j,M_j\thicksim \text{Beta}(\mu_j,M_j)
\end{align}

@@

The second population moment centered around mean (variance) is
\begin{equation}
 Var(\theta_{ji}) = \frac{\mu_j(1-\mu_j)}{M_j+1}
\end{equation}
while the sample variance is
\begin{equation}
 \begin{split}
  Var(\tilde{\theta}_{ji}) &= \frac{1}{N-1}\sum_{i=1}^N(\tilde{\theta}_{ji}-\text{E}[\tilde{\theta}_{ji}])^2\\ &= \frac{1}{N-1}\sum_{i=1}^N(\tilde{\theta}_{ji}-\tilde{\mu}_j)^2
 \end{split}
\end{equation}
By definition of Method of moments,
\begin{equation}
  \frac{\tilde{\mu}_j(1-\tilde{\mu}_j)}{\tilde{M}_j+1} = \frac{1}{N-1}\sum_{i=1}^N(\tilde{\theta}_{ji}-\tilde{\mu}_j)^2,
\end{equation}
so,
\begin{equation}
\begin{split}
 \tilde{M}_j &= (N-1)\cdot\frac{\tilde{\mu}_j(1-\tilde{\mu}_j)}{\sum_{i=1}^N(\tilde{\theta}_{ji}-\tilde{\mu}_j)^2}-1\\ &= (N-1)\cdot\frac{\left(\frac{1}{N-1}\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\right)\cdot\left(1-\frac{1}{N-1}\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\right)}{\sum_{i=1}^N\left(\frac{r_{ji}}{n_{ji}}-\frac{1}{N-1}\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\right)^2}\\ &= \frac{\left(\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\right)\cdot\left(1-\frac{1}{N-1}\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\right)}{\sum_{i=1}^N\left(\frac{r_{ji}}{n_{ji}}-\frac{1}{N-1}\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\right)^2}
\end{split}
\end{equation}

\subsubsection{$\mu_0$ and $M_0$}\

We set a hyperprior distribution $Beta(\mu_0,M_0)$ on $\mu_j$ based on our belief that $\mu_j\in[0,1]$
\begin{align}
 \mu_j|\mu_0,M_0\thicksim \text{Beta}(\mu_0,M_0).
\end{align}

@@

Using Methond of Moment
\begin{equation}
  \tilde{\mu}_0 = \text{E}[\tilde{\mu}_j] = \frac{1}{(J-1)(N-1)}\sum_{i=1}^J\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}
\end{equation}

The second population moment centered around mean (variance) is
\begin{equation}
 Var(\mu_j) = \frac{\mu_0(1-\mu_0)}{M_0+1}
\end{equation}
while the sample variance is
\begin{equation}
 \begin{split}
  Var(\tilde{\mu}_j) &= \frac{1}{J-1}\sum_{i=1}^J(\tilde{\mu}_j-\text{E}[\tilde{\mu}_j])^2\\ &= \frac{1}{J-1}\sum_{i=1}^J(\tilde{\mu}_j-\tilde{\mu}_0)^2
 \end{split}
\end{equation}
By definition of Method of moments,
\begin{equation}
  \frac{\tilde{\mu}_0(1-\tilde{\mu}_0)}{\tilde{M}_0+1} = \frac{1}{J-1}\sum_{i=1}^J(\tilde{\mu}_j-\tilde{\mu}_0)^2,
\end{equation}
so,
\begin{equation}
\begin{split}
 \tilde{M}_0 &= (J-1)\cdot\frac{\tilde{\mu}_0(1-\tilde{\mu}_0)}{\sum_{i=1}^J(\tilde{\mu}_j-\tilde{\mu}_0)^2}-1\\ &= (J-1)\cdot\frac{\left(\frac{1}{(J-1)(N-1)}\sum_{i=1}^J\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\right)\cdot\left(1-\frac{1}{(J-1)(N-1)}\sum_{i=1}^J\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\right)}{\sum_{i=1}^J\left(\frac{1}{N-1}\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}-\frac{1}{(J-1)(N-1)}\sum_{i=1}^J\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\right)^2}-1\\ &= \frac{\left(\sum_{i=1}^J\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\right)\cdot\left(1-\frac{1}{(J-1)(N-1)}\sum_{i=1}^J\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\right)}{\sum_{i=1}^J\left(\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}-\frac{1}{J-1}\sum_{i=1}^J\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\right)^2}-1
\end{split}
\end{equation}
\bibliographystyle{apalike}
\bibliography{bioinfo}
\end{document}
